---
title: "Probability"
subtitle: "Intro to Data Science"
author: Anton Kalén
institute: University of Skövde
menu: false
date: 2022-09-16
date-format: medium
format:
  revealjs:
    echo: true
    auto-animate-duration: 0.3
code-link: true
self-contained: true
bibliography: references.bib
---

> Mathematics is the logic of certainty;
>
> probability is the logic of uncertainty.

## Certainty {auto-animate="true"}

::: {.callout-note appearance="simple" icon="false"}
## Constants

$\pi = 3.1415927$
:::

. . .

```{r}
#| output-location: column
x <- seq(
  from = 1, 
  to = 100, 
  length.out = 100
)

pi <- rep(
  3.1415927, 
  times = 100
)
```

## Certainty {auto-animate="true"}

::: {.callout-note appearance="simple" icon="false"}
## Constants

$\pi = 3.1415927$
:::

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"

x <- seq(
  from = 1, 
  to = 100, 
  length.out = 100
)

pi <- rep(
  3.1415927, 
  times = 100
)

par(mar = c(5.1, 4.1, 0.01, 0.01))
plot(x, pi, type = "l")
```

## Certainty {auto-animate="true"}

::: {.callout-note appearance="simple" icon="false"}
### Deterministic variables:

$VAT = 0.25 \times price$
:::

. . .

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
#| 
price <- seq(
  from = 1, 
  to = 100, 
  by = 1
)

vat = 0.25 * price
```

## Certainty {auto-animate="true"}

::: {.callout-note appearance="simple" icon="false"}
## Deterministic variables

$VAT = 0.25 \times price$
:::

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"

price <- seq(
  from = 1, 
  to = 100, 
  by = 1
)

vat = 0.25 * price

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(price, vat, type = "l")
```

## Uncertainty

Will this person buy or not?

. . .

Number of visitors tomorrow

. . .

Height of the next person coming

. . .

Total weight of the next transport

## Exercise

> You work in a online retailer and are tasked with predicting if a visitor will buy the product sold or not.

In groups of 2--4 persons, discuss what sources of uncertainty exist that makes it hard to be sure about if the visitor buys or not.
5 min.

## Sources of Uncertainty

-   Lack of knowledge of what determines if a visitor buys.
-   Lack of knowledge about the visitor.
-   Measurement error of potential predictors.
-   Measurement error of previous sales.
-   Sampling uncertainty of past visitors.
-   Randomness.

## Probability of buys {auto-animate="true"}

> The retailer has 3 visitors per day.
> How many of them will buy a product?

## Probability of buys {auto-animate="true"}

1.  What are all possible outcomes?

::: {.callout-note appearance="simple" icon="false"}
### Sample space

$\Omega = \{\text{nnn}, \text{bnn}, \text{nbn}, \text{nnb}, \text{bbn}, \text{bnb}, \text{nbb}, \text{bbb} \}$

$\text{n} = \text{no buy}$

$\text{b} = \text{buy}$
:::

## Probability of buys {auto-animate="true"}

2.  What are we interested in knowing?

::: {.callout-note appearance="simple" icon="false"}
### Random Variable

Let $X$ be the number of buys in a day.

$X : \Omega \to \mathbb{R}$
:::

::: aside
A random variable is a function mapping the sample space, $\Omega$, to the real line, $\mathbb{R}$ (i.e., it assigns a number to each possible event).
:::

## Probability of buys {auto-animate="true"}

2.  What are we interested in knowing?

::: {.callout-note appearance="simple" icon="false"}
### Random Variable

$X(\text{nnn}) = 0$

$X(\text{bnn}) = 1$

$X(\text{nbn}) = 1$

$X(\text{nnb}) = 1$

$X(\text{bbn}) = 2$

$X(\text{bnb}) = 2$

$X(\text{nbb}) = 2$

$X(\text{bbb}) = 3$
:::

## Probability of buys {auto-animate="true"}

3.  What is the probability of the different outcomes?

::: {.callout-note appearance="simple" icon="false"}
## Probability function

$P(X)$

$P : X \to [0, 1]$
:::

::: aside
The probability function, $P$, assigns a probability between 0 and 1 to each possible outcome of the random variable $X$
:::

## Probability of buys {auto-animate="true"}

3.  What is the probability of the different outcomes?

::: {.callout-note appearance="simple" icon="false"}
## Probability function

$P(X = 0) = \frac{1}{8} = 0.125$

$P(X = 1) = \frac{3}{8} = 0.375$

$P(X = 2) = \frac{3}{8} = 0.375$

$P(X = 3) = \frac{1}{8} = 0.125$
:::

::: {.callout-caution appearance="simple" icon="false"}
If probability of each visitor buying = 0.5
:::

## Probability notation

$X$ --- Random variable (r.v.)

$\Omega_X = \{x_1, \ldots, x_n \}$ --- Sample space of r.v.
$X$

$P(X)$ --- Probability function on r.v.
$X$

$P(X = x_1)$ or $P(x_1)$ --- Probability of $X$ taking value $x_1$

## Probability distribution

We can represent random variables using probability distributions.

::: {.callout-note appearance="simple" icon="false"}
## Distribution of r.v.

$X \sim \text{Bin}(n, p)$

$n = \text{number of trials}$.

$p = \text{probability of success}$.
:::

::: aside
The random variable $X$ is binomial distributed, with $n$ trials and $p$ probability of success.
:::

## Binomial distribution

> The retailer has 3 visitors per day.
> The probability of each visitor buying a product is 0.5.
> How many of them will buy a product?

. . .

$n = 3$

$p = 0.5$

$X \sim \text{Bin}(3, 0.5)$

## Binomial distribution

> The retailer has **3** visitors per day.
> The probability of each visitor buying a product is **0.5**.
> How many of them will buy a product?

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
n <- 3
p <- 0.5

omega <- 0:n

prob <- dbinom(
  omega, 
  size = n, 
  prob = p
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(omega, prob, type = "h")
points(omega, prob, pch = 16)

```

## Binomial distribution

> The retailer has **10** visitors per day.
> The probability of each visitor buying a product is **0.5**.
> How many of them will buy a product?

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
#| code-line-numbers: "1"
n <- 10
p <- 0.5

omega <- 0:n

prob <- dbinom(
  omega, 
  size = n, 
  prob = p
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(omega, prob, type = "h")
points(omega, prob, pch = 16)

```

## Binomial distribution

> The retailer has **10** visitors per day.
> The probability of each visitor buying a product is **0.3**.
> How many of them will buy a product?

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
#| code-line-numbers: "2"
n <- 10
p <- 0.3

omega <- 0:n

prob <- dbinom(
  omega, 
  size = n, 
  prob = p
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(omega, prob, type = "h")
points(omega, prob, pch = 16)

```

## 

::: {.callout-note appearance="simple" icon="false"}
## Probability mass function (PMS)

$p_X(x_i) = P(X = x_i)$

`dbinom()`
:::

::: aside
The probability that the discrete r.v.
$X$ is equal to $x_i$.
:::

## 

::: {.callout-note appearance="simple" icon="false"}
## Probability mass function (PMS)

$p_X(x_i) = P(X = x_i)$

`dbinom()`
:::

::: {.callout-note appearance="simple" icon="false"}
## Cumulative distribution function (CDF)

$F_X(x_i) = P(X \leq x_i)$

`pbinom()`
:::

::: aside
The probability that the discrete r.v.
$X$ is smaller or equal to $x_i$.
:::

## Binomial function

::: columns
::: {.column width="50%"}
```{r}
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
#| echo: false
n <- 10
p <- 0.5

omega <- 0:n

prob <- dbinom(
  omega, 
  size = n, 
  prob = p
)

par(mar = c(5.1, 4.1, 4.1, 0.01))

plot(omega, prob, type = "h", main = "Binomial PMS (n = 10, p = 0.5)")
points(omega, prob, pch = 16)
```
:::

::: {.column width="50%"}
```{r}
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
#| echo: false
n <- 10
p <- 0.5

omega <- 0:n

prob <- pbinom(
  omega, 
  size = n, 
  prob = p
)

par(mar = c(5.1, 4.1, 4.1, 0.01))

plot(omega, prob, type="s", main = "Binomial CDF (n = 10, p = 0.5)")
```
:::
:::

## Poisson distribution

> Discrete duistribution for counting success in a specific time, where there is a large number of events and low probability.

. . .

For example, the online retailer has thousands of visitors each day and only a small proportion ends up buying.

. . .

::: {.callout-note appearance="simple" icon="false"}
## Poisson distribution

$X \sim \text{Pois}(\lambda)$

$\Omega_X = \mathbb{I}^+$ (Possible values are positive integers)

$\lambda$ --- Mean and variance of $X$
:::

## Poisson distribution (PMF)

> The average amount of daily sales are **4**.

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
lambda <- 4

range <- 0:30

prob <- dpois(
  range, 
  lambda = lambda,
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(range, prob, type = "h")
points(range, prob, pch = 16)

```

## Poisson distribution (PMF)

> The average amount of daily sales are **10**.

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
lambda <- 10

range <- 0:30

prob <- dpois(
  range, 
  lambda = lambda,
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(range, prob, type = "h")
points(range, prob, pch = 16)

```

## Poisson distribution (CDF)

> The average amount of daily sales are **4**.

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
lambda <- 4

range <- 0:30

prob <- ppois(
  range, 
  lambda = lambda,
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(range, prob, type = "s")
```

## Poisson distribution (CDF)

> The average amount of daily sales are **4**.

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
lambda <- 10

range <- 0:30

prob <- ppois(
  range, 
  lambda = lambda,
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(range, prob, type = "s")
```

## Conditional probability

> What is the probability that a visitor buys a product, given that it is a raining day?

. . .

|      | Buy | No Buy |
|------|----:|-------:|
| Sun  |  13 |     39 |
| Rain |  26 |    104 |

. . .

::: {.callout-note appearance="simple" icon="false"}
## Random variables & sample space

Finish visit: $X$, $\Omega_X = \{\text{Buy}, \text{No Buy}\}$

Wheather: $Y$, $\Omega_Y = \{\text{Sun}, \text{Rain}\}$
:::

## Conditional probability

> What is the probability that a visitor buys a product, given that it is a raining day?

|      | Buy | No Buy |
|------|----:|-------:|
| Sun  |  13 |     39 |
| Rain |  26 |    104 |

::: {.callout-note appearance="simple" icon="false"}
## Marginal probability

Marginal probabilities $X$: $P(X = \text{Buy})$, $P(X = \text{No buy})$

Marginal probabilities of $Y$: $P(Y = \text{Sun})$, $P(Y = \text{Rain})$
:::

## Conditional probability

> What is the probability that a visitor buys a product, given that it is a raining day?

|      | Buy | No Buy |
|------|----:|-------:|
| Sun  |  13 |     39 |
| Rain |  26 |    104 |

::: {.callout-note appearance="simple" icon="false"}
## Conditional probability $P(X|Y)$

Probability of $X$ given Sun: $P(X|Y = \text{Sun})$, or $P(X|\text{Sun})$

Probability of Buy, given Rain: $P(X = \text{Buy}| Y = \text{Rain})$, or $P(\text{Buy}|\text{Rain})$
:::

## Conditional probability

> What is the probability that a visitor buys a product, given that it is a raining day?

|      | Buy | No Buy |
|------|----:|-------:|
| Sun  |  13 |     39 |
| Rain |  26 |    104 |

::: {.callout-note appearance="simple" icon="false"}
## Joint probability $P(X, Y)$

Probability of Buy when it rains: $P(X = \text{Buy}, Y = Rain)$, or $P(X = \text{Buy}, Y = Rain)$
:::

## Conditional probability

```{r}
# Create the matrix

sun <- c(13, 39)
rain <- c(26, 104)

sales <- rbind(sun, rain)

colnames(sales) <- c("buy", "no buy")

sales
```

## Conditional probability

Marginal probability $P(X = \text{Buy})$

. . .

```{r}
# P(X = Buy)
sum(sales[,"buy"]) / sum(sales)
```

. . .

Marginal probability $P(Y = \text{Sun})$

. . .

```{r}
# P(Y = Sun)
sum(sales["sun",]) / sum(sales)
```

## Conditional probability

Conditional probability $P(X = \text{Buy}|Y = \text{Rain})$

. . .

```{r}
# P(X = Buy|Y = Rain)
sales["rain","buy"] / sum(sales["rain", ])
```

. . .

Conditional probability $P(X = \text{Buy}|Y = \text{Sun})$

. . .

```{r}
# P(X = Buy|Y = Sun)
sales["sun","buy"] / sum(sales["sun", ])
```

. . .

Conditional probability $P(X|Y)$

```{r}
sales / rowSums(sales)
```

## Conditional probability

Joint probability $P(X = \text{Buy}, Y = \text{Rain})$

```{r}
sales["rain", "buy"] / sum(sales)
```

. . .

Joint probability $P(X, Y)$

```{r}
sales / sum(sales)
```

. . .

Probabilities sums to 1

```{r}
sum(sales / sum(sales))
```

## Continuous random variables

. . .

> Random variables that can take any real value in an interval

-   Infinite amount of real numbers → any specific real number has zero probability.

. . .

-   Probability can only exist over measurable areas/volumes.

. . .

-   Instead of probability mass functions (PMF), we work with probability densities functions (PDF)

## Normal distribution

::: {.callout-note appearance="simple" icon="false"}
$X \sim \mathcal{N}(\mu, \sigma)$

$mu$ --- Mean of $X$

$\sigma$ --- Standard deviation of $X$

$\Omega = \mathbb{R} = (-\infty, \infty)$
:::

## Normal distribution

::: {.callout-note appearance="simple" icon="false"}
## Probability density function

$\displaystyle f(X)={\frac {1}{\sigma {\sqrt {2\pi }}}}e^{-{\frac {1}{2}}\left({\frac {x-\mu }{\sigma }}\right)^{2}}$

$\displaystyle \int_{-\infty}^{\infty} f(X) dX = 1$
:::

::: aside
The probability density function $f(X)$ over all real numbers integrates to one .
:::

## Normal distribution

::: {.callout-note appearance="simple" icon="false"}
## Cumulative distribution function (CDF)

$\displaystyle F(X) = \int_{-\infty}^{X} f(X) dX$
:::

## Normal distribution

Probability density function of $\mathcal{N}(0, 1)$.

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
mu <- 0
sigma <- 1

x <- seq(-10, 10, by = .1)

prob <- dnorm(
  x, 
  mean = mu, 
  sd = sigma
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(x, prob, type = "l")
```

## Normal distribution

Probability density function of $\mathcal{N}(0, 2)$.

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
#| code-line-numbers: "2"
mu <- 0
sigma <- 2

x <- seq(-10, 10, by = .1)

prob <- dnorm(
  x, 
  mean = mu, 
  sd = sigma
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(x, prob, type = "l")
```

## Normal distribution

Probability density function of $\mathcal{N}(3, 2)$.

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
#| code-line-numbers: "1,2"
mu <- 3
sigma <- 2

x <- seq(-10, 10, by = .1)

prob <- dnorm(
  x, 
  mean = mu, 
  sd = sigma
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(x, prob, type = "l")
```

## Normal distribution

Cumulative distribution function of $\mathcal{N}(0, 1)$.

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
#| code-line-numbers: "6,7,8,9,10"
mu <- 0
sigma <- 1

x <- seq(-10, 10, by = .1)

prob <- pnorm(
  x, 
  mean = mu, 
  sd = sigma
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(x, prob, type = "l")
```

## Normal distribution

Cumulative distribution function of $\mathcal{N}(0, 2)$.

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
#| code-line-numbers: "2"
mu <- 0
sigma <- 2

x <- seq(-10, 10, by = .1)

prob <- pnorm(
  x, 
  mean = mu, 
  sd = sigma
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(x, prob, type = "l")
```

## Normal distribution

Cumulative distribution function of $\mathcal{N}(3, 2)$.

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
#| code-line-numbers: "1"
mu <- 3
sigma <- 2

x <- seq(-10, 10, by = .1)

prob <- pnorm(
  x, 
  mean = mu, 
  sd = sigma
)

par(mar = c(5.1, 4.1, 0.01, 0.01))

plot(x, prob, type = "l")
```

## Normal distribution

Given $X \sim \mathcal{N}(1, 0)$, the $P(X \leq 1) = F_X(1)$.

```{r}
mu <- 0
sigma <- 1
x <- 1

pnorm(
  x, 
  mean = mu, 
  sd = sigma
)
```

## Normal distribution

Given $X \sim \mathcal{N}(1, 0)$, calculate the probability that $x \in [-1, 1]$.

```{r}
mu <- 0
sigma <- 1
x_1 <- -1
x_2 <- 1

p_1 <- pnorm(x_1, mu, sigma)

p_2 <- pnorm(x_2, mu, sigma)

p_2 - p_1
```

## Normal distribution

Simulate 1000 draws from $X \sim \mathcal{N}(1, 0)$

```{r}
#| output-location: column
#| fig-asp: 0.9
#| fig-width: 4
#| out-width: "90%"
mu <- 0
sigma <- 1
n <- 1000

draws <- rnorm(n, mu, sigma)

par(mar = c(5.1, 4.1, 4.1, 0.01))

hist(draws)
```

# Lab assignment

## Lab assignment

-   Similar structure to Lab 1: R programming

-   You need to hand in correct solution with working code for **all** excercises.

-   Work individual:

    -   One report per student.
    -   No duplicate reports.

-   You can only use base R, no external libraries.

-   No copy paste from internet or other sources.

## Lab assignment

-   You upload a PDF report in Canvas.
-   PDF should contain:
    -   Text explanation of your solution (choices etc.)
    -   R code (carefully commented)
    -   Any plots/figures should be included

## Lab assignment

-   Make sure code is easy to test the code by copy-pasting it into an R-session.
-   Reports will be checked for plagarism.
-   Two opportunities to submit.

## Resources

-   Introduction to Probability (free online book): <http://probabilitybook.net>
-   Harvard's "Statistics 110: Probability"-lectures: <https://projects.iq.harvard.edu/stat110/youtube>
-   Base R cheat sheet: <https://github.com/rstudio/cheatsheets/blob/main/base-r.pdf>
-   R help pages: `?distributions`, `?plot` ...