\documentclass{article}
\usepackage{fancyhdr} 
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{enumitem}
\fancyhead[R]{Probability -- Lab Assignment}
\pagestyle{fancy}

\author{
        \large
        \textsc{Anton Kalén} \\
        \texttt{anton.kalen@his.se}
}                                      

\title{Probability -- Lab Assignment
{\large {\bf IMPORTANT:} See further instructions regarding the report and work structure from the R lecture slides. In particular, you should work  individually.}           
\\[1cm]}
\date{\today}

\begin{document}
\maketitle
\thispagestyle{empty}   
\begin{enumerate}
	\item Calculate the probability that $x \in [0, 3]$ for a standard 
		normal distribution ($\mu = 1$, $\sigma = 2$).	
	\item Draw different number of observations from the standard normal 
		distribution ($\mathcal{N}(0,1)$), e.g., 10, 30, 50 and 100, and calculate the mean value.
		What happens as the number of observations increases?
	\item For a random variable $X \sim \text{Binom}(0.7, 10)$, plot its probability mass function.
	\item For the above distribution, calculate central tendencies (mean, median, mode). Are they similiar? Discuss why/why not.
	\item In what is known as hierarchical models or multi-level models, one assumes distributions over parameters in
		other distributions, i.e., a hierarchy. Assume that there is a 
		standard normal distribution over $\mu$ in another standard normal
		distribution and generate observations $x$ based on these assumptions.
		\begin{equation}
		\begin{split}
			x &\sim N(\mu, \sigma=1)\\
			\mu &\sim N(0,1)
		\end{split}
		\end{equation}
		Plot the drawn samples in a histogram.
	\item Read up on the exponential distribution. Explain what an exponential random variable likely represent (what is it used for). Plot its probability density function.
	\item You work for an amusement park that wants to decisde how many people they have one person selling tickets in the entrance. On average, they sell 4 tickets per minute, but it varies from minute to minute. What is a reasonable distribution to assume for the random variable $X$ indicating the amount of tickets per minute? Plot its probability mass/density function, as well as its cumulative distribution function.
	\item What is the probability of the above amusement park selling more than 8 tickets one minute?
\end{enumerate}
\end{document}

